Using high-resolution variant frequencies to empower clinical genome interpretation

### 2016-10-11 manuscript/manuscript.Rmd
Wording changes

Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts. We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants. We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined thresholds, which are now available through the ExAC data browser and for download, to assist others in applying our framework.

@@

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. Variant penetrance of 0.5 was used for all analyses, as penetrance estimates for individual variants are not widely available. This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders. When the population frequency of a variant is known, we can use this to estimate prevalence `r citep("10.1126/scitranslmed.aad5169")`.

@@

#### *Application and validation in hypertrophic cardiomyopathy*

We illustrate our general approach using the dominant cardiac disorder hypertrophic cardiomyopathy (HCM), which has an estimated prevalence of 1 in 500 in the general population `r citep("10.1038/gim.2014.205")`. As there have been previous large-scale genetic studies of HCM, with series of up to 6,179 individuals `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, we can make the assumption that no newly identified variant will be more frequent in cases that those identified to date (at least for well-studied ancestries). This allows us to define the maximum allelic contribution of any single variant to the disorder. In these large case series, the largest proportion of cases is attributable to the missense variant _MYBPC3_ c.1504C>T (p.Arg502Trp), found in 104/6179 HCM cases (1.7%; 95%CI 1.4-2.0%) `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. We therefore take the upper bound of this proportion (0.02) as an estimate of the maximum allelic contribution in HCM (Table 1). Our maximum expected population allele frequency for this allele, assuming 50% penetrance as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1). 
To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population allele frequency of 4.0x10^-5^ can be observed in a random population sample of a given size. For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population allele frequency (**Methods**). For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes). The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^; Table 1).

@@

We computed the maximum tolerated allele count for four additional cardiac disorders with available cases series. In all cases the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).

@@

To assess the HCM thresholds empirically, we explored the ExAC allele frequency spectrum of `r myVariantsN` distinct autosomal variants identified in 6179 recently published HCM cases referred for diagnostic sequencing, and individually assessed and reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. 
`r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases. `r myAbsentPath` of these variants are absent from ExAC.  The `r myFailPath` variants historically classified as ‘Likely Pathogenic’, but prevalent in ExAC in this analysis, were reassessed using contemporary ACMG criteria: there was no strong evidence in support of pathogenicity, and they were reclassified in light of these findings (Table S1). This analysis identifies `r myFailVus`/`r myAllVus` (`r signif(100*myFailVus/myAllVus,3)`%) VUS that are unlikely to be truly causative for HCM.

@@

To further validate this approach, we examined all 601 variants identified in ClinVar `r citep("10.1093/nar/gkt1113")` as "Pathogenic" or "Likely Pathogenic" and non-conflicted for HCM. 558 (93%) were sufficiently rare when assessed as described. 43 variants were insufficiently rare in at least one ExAC population, and were therefore re-curated. 42 of these had no segregation or functional data sufficient to demonstrate pathogenicity in the heterozygous state, and we would classify as VUS at most. The remaining variant (MYBPC3:c.3330+5G>C) had convincing evidence of pathogenicity, though with uncertain penetrance (see **Supplementary Methods**), and was observed twice in the African/African American ExAC population. This fell outside the 95% confidence interval for an underlying population frequency <4x10^-5^, but within the 99% confidence threshold: a single outlier due to stochastic variation is unsurprising given that these nominal probabilities are not corrected for multiple testing across 601 variants. In light of our updated assessment, 20 variants were reclassified as Benign/Likely Benign and 22 as VUS according to the American College for Medical Genetics and Genomics (ACMG) guidelines for variant interpretation `r citep("10.1038/gim.2015.30")` (Table S1).

After curating variants above our calculated HCM threshold, the false discovery rate was 0/477 (0.000; 95%CI 0.000-0.008) and 1/559 (0.002; 95%CI 0.000-0.010) for the published HCM cohort and ClinVar data respectively.

@@

Intuitively, if one penetrant recessive variant were to be responsible for all PCD cases, it could have a maximum population frequency of $\sqrt(1/10000)$. The maximum frequency of a recessive disease-causing variant in the population can be more completely defined as:  
*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$

where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant (full derivation can be found in **Methods**).

We can refine our evaluation of PCD by estimating the maximum genetic and allelic contribution. Across previously published cohorts of PCD cases `r citep(c("10.1164/rccm.200603-370OC","10.1164/rccm.200601-084OC","10.1038/ng.2277"))`, *DNAI1* IVS1+2\_3insT was the most common variant with a total of 17/358 alleles (4.7% 95CI 2.5-7.0%). Given that ~9% of all patients with PCD have disease-causing variants in _DNAI1_ and the IVS1+2\_3insT variant is estimated to account for ~57% of variant alleles in _DNAI1_ `r citep("10.1164/rccm.200603-370OC")`, we can take these values as estimates of the maximum genetic and allelic contribution for PCD, yielding a maximum expected population AF of $\sqrt(1/10000) \times 0.57 \times \sqrt0.09 \times 1/\sqrt0.5 = 2.42 \times10$^-3^ This translates to a maximum tolerated ExAC AC of 322. *DNAI1* IVS1+2\_3insT is itself present at 56/121108 ExAC alleles (45/66636 non-Finnish European alleles). A single variant reported to cause PCD in ClinVar occurs in ExAC with AC > 332 (_NME8_ NM\_016616.4:c.271-27C>T; AC=2306/120984): our model therefore indicates that this variant frequency is too common to be disease-causing, and consistent with this we note that it meets none of the current ACMG criteria for assertions of pathogenicity, and have reclassified it as VUS (see **Supplementary Methods**).

@@

For each ExAC variant, we defined a "filtering allele frequency" that represents the threshold disease-specific "maximum credible allele frequency" at or below which the disease could not plausibly be caused by that variant. A variant with a filtering allele frequency≥the user-defined maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate (Figure 3). The filtering allele frequency is not desease speific and hence can be applied to any disease of interest. This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).

@@

### Data availability

@@

We recognise several limitations of our approach. First, the approach is limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters. In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations. It is critical to define the genetic architecture in the population under study. We acknowledge that there is no substitute for knowning the genetic architecture of the condition of interest, and for some conditions this is not possible. It is the nature of any framework to rely on underlying assumptions, and we believe our method makes these explicit. If a parameter is particularly unclear, it is possible to define a range of architectures that fit with the observed variant. Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants (see **Supplementary Methods** for alternative approaches).

@@

>>>>

Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts. We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants. We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined disease-specific thresholds, which are now available through the ExAC data browser and for download, to assist others in applying our framework.

@@

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. For cardiovascular diseases variants were modelled with a pentrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found accross a range of variants/disorders.

@@

#### *Application and validation in hypertrophic cardiomyopathy*
#### *Defining maximum credible population allele frequency*

We illustrate our general approach using the dominant cardiac disorder hypertrophic cardiomyopathy (HCM), which has an estimated prevalence of 1 in 500 in the general population `r citep("10.1038/gim.2014.205")`. As there have been previous large-scale genetic studies of HCM, with series of up to 6,179 individuals `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, we can make the assumption that no newly identified variant will be more frequent in cases that those identified to date (at least for well-studied ancestries). This allows us to define the maximum allelic contribution of any single variant to the disorder. In these large case series, the largest proportion of cases is attributable to the missense variant _MYBPC3_ c.1504C>T (p.Arg502Trp), found in 104/6179 HCM cases (1.7%; 95%CI 1.4-2.0%) `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. We therefore take the upper bound of this proportion (0.02) as an estimate of the maximum allelic contribution in HCM (Table 1). Our maximum expected population allele frequency for this allele, assuming 50% penetrance as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1). 
#### *Controlling for sample variation*
To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population allele frequency of 4.0x10^-5^ can be observed in a random population sample of a given size. For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population allele frequency (**Methods**). For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes). The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^; Table 1).

@@

#### *Assessing the acuracy of our approach*
For all diseases with cases series that permitted us to define the genetic arcitecture, the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).

@@

To assess the HCM thresholds empirically, we explored the ExAC allele frequency spectrum of `r myVariantsN` distinct autosomal variants identified in 6179 recently published HCM cases referred for diagnostic sequencing, and individually assessed and reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. 
`r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases. The `r myFailPath` variants historically classified as ‘Likely Pathogenic’, but prevalent in ExAC in this analysis, were reassessed using contemporary ACMG criteria: there was no strong evidence in support of pathogenicity, and they were reclassified in light of these findings (Table S1). This analysis identifies `r myFailVus`/`r myAllVus` (`r signif(100*myFailVus/myAllVus,3)`%) VUS that are unlikely to be truly causative for HCM.

@@

To further validate this approach, we examined all 601 variants identified in ClinVar `r citep("10.1093/nar/gkt1113")` as "Pathogenic" or "Likely Pathogenic" and non-conflicted for HCM. 558 (93%) were sufficiently rare when assessed as described. 43 variants were insufficiently rare in at least one ExAC population, and were therefore re-curated. 42 of these had no segregation or functional data sufficient to demonstrate pathogenicity in the heterozygous state, and would be classified by the ACMG framework as VUS at most. The remaining variant (MYBPC3:c.3330+5G>C) had convincing evidence of pathogenicity, though with uncertain penetrance (see **Supplementary Methods**), and was observed twice in the African/African American ExAC population. This fell outside the 95% confidence interval for an underlying population frequency <4x10^-5^, but within the 99% confidence threshold: a single outlier due to stochastic variation is unsurprising given that these nominal probabilities are not corrected for multiple testing across 601 variants. In light of our updated assessment, 20 variants were reclassified as Benign/Likely Benign and 22 as VUS according to the American College for Medical Genetics and Genomics (ACMG) guidelines for variant interpretation `r citep("10.1038/gim.2015.30")` (Table S1).

After curating variants above our calculated HCM threshold, the false positive rate was 0/477 (0.000; 95%CI 0.000-0.008) and 1/559 (0.002; 95%CI 0.000-0.010) for the published HCM cohort and ClinVar data respectively.

@@

Intuitively, if one penetrant recessive variant were to be responsible for all PCD cases, it could have a maximum population frequency of $\sqrt(1/10000)$. We can refine our evaluation of PCD by estimating the maximum genetic and allelic contribution (see *Methods*). Across previously published cohorts of PCD cases `r citep(c("10.1164/rccm.200603-370OC","10.1164/rccm.200601-084OC","10.1038/ng.2277"))`, *DNAI1* IVS1+2\_3insT was the most common variant with a total of 17/358 alleles (4.7% 95CI 2.5-7.0%). Given that ~9% of all patients with PCD have disease-causing variants in _DNAI1_ and the IVS1+2\_3insT variant is estimated to account for ~57% of variant alleles in _DNAI1_ `r citep("10.1164/rccm.200603-370OC")`, we can take these values as estimates of the maximum genetic and allelic contribution for PCD, yielding a maximum expected population AF of $\sqrt(1/10000) \times 0.57 \times \sqrt0.09 \times 1/\sqrt0.5 = 2.42 \times10$^-3^ This translates to a maximum tolerated ExAC AC of 322. *DNAI1* IVS1+2\_3insT is itself present at 56/121108 ExAC alleles (45/66636 non-Finnish European alleles). A single variant reported to cause PCD in ClinVar occurs in ExAC with AC > 332 (_NME8_ NM\_016616.4:c.271-27C>T; AC=2306/120984): our model therefore indicates that this variant frequency is too common to be disease-causing, and consistent with this we note that it meets none of the current ACMG criteria for assertions of pathogenicity, and have reclassified it as VUS (see **Supplementary Methods**).

@@

For each ExAC variant, we defined a "filtering allele frequency" that represents the threshold disease-specific "maximum credible allele frequency" at or below which the disease could not plausibly be caused by that variant. A variant with a filtering allele frequency≥the maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate. The filtering allele frequency is not desease speific and hence can be applied to any disease of interest by comparing with a user-defined disease-specific maximum credible allele frequency (Figure 3). This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).

@@

@@

We recognise several limitations of our approach. First, we are limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters. In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations. Unfortunately, there is no substitute for understanding the genetic architecture of the disease of interest, but when considering a specific variant that has been observed at low frequency in a given reference population, our framework and accompanying web tool provide a transparent approach to define the range of disease architectures that are compatible with the observed data. For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value.

Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants. Although we would argue that variants with low penetrance have questionable diagnostic utility, our calculator app allows a user to define a range of compatible penetrance for a given allele frequency (see **Supplementary Methods**). Furthermore, methods to estimate variant penetrance from large case and control cohorts have been described previously`r citep("10.1126/scitranslmed.aad5169")`.

@@

## DATA AVAILABILITY
